An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2022-07-07 2022/ICNAAM/ICNAAM.tex

% Title portion
\title{Using Machine Learning for Subdivision of Parameters in Global Optimization Problems of Large Dimensionality}

@@

At present, the global optimization methods are applied to solving various problems arising in various fields of science and technology. 
For example, the evaluation of the model parameters from the experimental data is a traditional field of application of the global optimization methods.
Among these ones, there are, for example, the problems of chemical kinetics \cite{Gubaydullin2021}.
In the problems of this kind, it is necessary to perform a search for the values of unknown model parameters, at which the modeling results are the closest to 
the ones obtained experimentally.

@@

As a rule, one can't subdivide the parameters into the groups with different character of the objective function behavior in advance since the objective function in applied problems is usually defined as a black box. 
So far, the development of the algorithm for the minimization of the essentially multiextremal functions taking into account different character of the dependence of the objective function on different groups of parameters is relevant. 
In the present work, a special method for such a subdivision based on the analysis of the regression model of the function with respect to each variable separately has been proposed.
To confirm the efficiency of the scheme proposed, a representative set of test problems has been solved.

@@

Let us assume the function $\varphi(y)$ to be a multiextremal one and to satisfy the Lipschitz condition
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty,
\]
with the constant $L$ unknown {\em a priori}.
At the same time, let us assume the dependence of the objective function on $N_1$ parameters to be multiextremal while the dependence on the rest $N_2 = N - N_1$ parameters is close to the linear or quadratic one.
At the same time, the subdivision of the parameters is unknown. 
In the present work, an approach to the subdivision of the problem parameters into the two specified groups is proposed.
The details of the method are described in the next section.

@@

In the present study, we used the global search algorithm combined with the dimensionality reduction scheme based on Peano curves \cite{Sergeyev2013,Lera2021} to solve the global optimization problem (\ref{global_problem}). 
This algorithm was demonstrated to be a powerful tool for solving complex multiextremal problems \cite{Kvasov2013,Kalyulin2017,Cavoretto2021}.
To solve the local problem (\ref{local_problem}), we used Hooke-Jeevse pattern search method (see, e.g., \cite{Kelley}). 
Unlike the BFGS-type methods (see, e.g., \cite{Nocedal}), this algorithm doesn't require computing the gradient and works well under conditions of the 
computing errors in the function values that is typical in solving the applied problems.

\section{SUBDIVISION OF THE PROBLEMS PARAMETERS}

@@

Step 1. Fix the reference point $\overline{y} = (\overline{y}_1, \overline{y}_2,...,\overline{y}_N)$ for the objective function $\varphi(y)$. The choice of the point is performed by one of the following methods:
\begin{enumerate}
\item the point is fixed according to the physical meaning of the applied problem being solved;
\item use the solution found by the local method in $K$ iterations (in the present study, Hooke-Jeevse pattern search method was used) as the reference point.
\item use the solution found by the global method in $K$ iterations (in the present study, the global search algorithm was used) as the reference point.
\end{enumerate}

Step 2. For all components of $y_i, \; 1\leq i \leq N$, investigate the locality:
\begin{enumerate}
\item compute $P+1$ values of the objective function $\varphi(y)$ at the points $z_i^j = (\overline{y}_1,...,\overline{y}_{i-1},y_i^j,\overline{y}_{i+1},...,\overline{y}_N)$ where $y_i^j =  a_i + jh, \; h=(b_i-a_i)/P, \; 0\leq j \leq P$;
\item form a set $Q$ comprising of $P+1$ pairs of the kind $\left\{y_i^j, \varphi(z_i^j)\right\} $ 
\item construct a regression model of given power $deg$ based on the data from $Q$;
\item compute the coefficient of determination $R^2$ and perform the classification of the $i^{\rm th}$ variable:
\begin{itemize}
\item if $R^2$ is greater than a predefined threshold $T$, then add the variable to the set of the local variables,
\item else add the variable to the set of the global variables.
\end{itemize}
\end{enumerate}

@@

The regression is based on the least squares method (LSM).
The quality of the model constructed was evaluated at Step 2, Item 4 using the adjusted coefficient of determination $R^2_a$.
This coefficient features the similarity of the regression model constructed to the objective function in the fixed section.
The value of $R^2_a$, as a rule, falls between 0 and 1.
A result close to 1 shows the accurasy of the model to be high.

@@

As a result of solving the Gr-L series with the preset precision of $0.01$ and the parameters of the method $K=100$, $deg=1$, and $T=0.6$, correct subdivision of he variables into the global and local ones were performed in all problems.
Then, all problems were solved successfully using the nested optimization scheme in appropriate time and number of iterations.
However, the solving of the complicated series Gr-LQ with the use of the linear regression didn't give good results because of the presence of the parabolic-type variables, for which the percentages of similarity to the linear model were extremely low.
In this connection, other parameters of the method $K=100$, $deg=2$ (the quadratic regression), and $T=0.95$ were selected to conduct the experiments.
As a result, all parameters of the problem were subdivided into the local and global ones correctly, and all problems were solved successfully.
% in appropriate time and number of iterations.

>>>>

% Title portion
\title{Using Machine Learning for Separation of Parameters in Global Optimization Problems of Large Dimensionality}

@@

At present, the global optimization methods are applied to solving various problems arising in various fields of science and technology \cite{Pinter2006}. 
For example, the evaluation of the model parameters from the experimental data is a traditional field of application of the global optimization methods.
Among these ones, there are the problems of chemical kinetics \cite{Gubaydullin2021}.
In the problems of this kind, it is necessary to perform a search for the values of unknown model parameters, at which the modeling results are the closest to 
the ones obtained experimentally.

@@

As a rule, one can't subdivide the parameters into the groups with different character of the objective function behavior in advance since the objective function in applied problems is usually defined as a black box. 
So far, the development of the algorithm for the minimization of the essentially multiextremal functions taking into account different character of the dependence of the objective function on different groups of parameters is relevant. 
In the present work, a special method for such a separation based on the analysis of the regression model of the function with respect to each variable separately has been proposed.
To confirm the efficiency of the scheme proposed, a representative set of test problems has been solved.

@@

Let us assume the function $\varphi(y)$ to be a multiextremal one and to satisfy the Lipschitz condition
\[
\left|\varphi(y')-\varphi(y'')\right|\leq L\left\|y'-y''\right\|,\; y',y'' \in D,\; 0<L<\infty,
\]
with the constant $L$ unknown {\em a priori}.
At the same time, let us assume the dependence of the objective function on $N_1$ parameters to be multiextremal while the dependence on the rest $N_2 = N - N_1$ parameters is close to the linear or quadratic one.
At the same time, the subdivision of the parameters is unknown. 
In the present work, an approach to the separation of the problem parameters into the two specified groups is proposed.
The details of the method are described in the next section.

@@

In the present study, we used the global search algorithm combined with the dimensionality reduction scheme based on Peano curves \cite{Sergeyev2013,Lera2021} to solve the global optimization problem (\ref{global_problem}). 
This algorithm was demonstrated to be a powerful tool for solving complex multiextremal problems \cite{Kvasov2013,Kalyulin2017,Cavoretto2021}.
To solve the local problem (\ref{local_problem}), we used Hooke-Jeeves (pattern search) method (see, e.g., \cite{Kelley}). 
Unlike the BFGS-type methods (see, e.g., \cite{Nocedal}), this algorithm doesn't require computing the gradient and works well under conditions of the 
computing errors in the function values that is typical in solving the applied problems.

\section{SEPARATION OF THE PROBLEMS PARAMETERS}

@@

Step 1. Fix the reference point $\overline{y} = (\overline{y}_1, \overline{y}_2,...,\overline{y}_N)$ for the objective function $\varphi(y)$. The choice of the point is performed by one of the following methods:
\begin{enumerate}
\item the point is fixed according to the physical meaning of the applied problem being solved;
\item use the solution found by the local method (e.g. Hooke-Jeeves method) in $K$ iterations as the reference point;
\item use the solution found by the global method (e.g. global search algorithm) in $K$ iterations as the reference point.
\end{enumerate}

Step 2. For all components $y_i, \; 1\leq i \leq N$, investigate the locality:
\begin{enumerate}
\item compute $P+1$ values of the objective function $\varphi(y)$ at the points $z_i^j = (\overline{y}_1,...,\overline{y}_{i-1},y_i^j,\overline{y}_{i+1},...,\overline{y}_N)$, where \mbox{$y_i^j = a_i + jh$},  $h=(b_i-a_i)/P, \; 0\leq j \leq P$;
\item form a set $Q$ comprising of $P+1$ pairs of the kind $\left\{y_i^j, \varphi(z_i^j)\right\} $ 
\item construct a regression model of given power $deg$ based on the data from $Q$;
\item compute the coefficient of determination $R^2$ and perform the classification of the $i^{\rm th}$ variable:
\begin{itemize}
\item if $R^2$ is greater than a predefined threshold $T$, then add the variable to the set of the local variables;
\item else add the variable to the set of the global variables.
\end{itemize}
\end{enumerate}

@@

The regression is based on the least squares method (LSM).
The quality of the model constructed was evaluated at Step 2, Item 4 using the adjusted coefficient of determination $R^2_a$.
This coefficient features the similarity of the regression model constructed to the objective function in the fixed section.
The value of $R^2_a$, as a rule, falls between 0 and 1.
A result close to 1 shows the accuracy of the model to be high.

@@

As a result of solving the Gr-L series with the preset precision of $0.01$ and the parameters of the method $K=100$, $deg=1$, and $T=0.6$, correct separation of the variables into the global and local ones were performed in all problems.
Then, all problems were solved successfully using the nested optimization scheme in appropriate time and number of iterations.
However, the solving of the complicated series Gr-LQ with the use of the linear regression didn't give good results because of the presence of the parabolic-type variables, for which the percentages of similarity to the linear model were extremely low.
In this connection, other parameters of the method $K=100$, $deg=2$ (the quadratic regression), and $T=0.95$ were selected to conduct the experiments.
As a result, all parameters of the problem were subdivided into the local and global ones correctly, and all problems were solved successfully.
% in appropriate time and number of iterations.

### 2022-07-07 2022/OPTIMA/paper.tex

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
%
\title{An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}}
%
\titlerunning{Algorithm for finding efficient solutions in multiobjective problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeny Kozinov\orcidID{0000-0001-6776-0096}}
%
\authorrunning{K. Barkalov, V. Grishagin and E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru}, \email{vagris@unn.ru}}

@@

In the first experiment 100 two-dimensional bi-criteria problems have been taken \textcolor{red}{($N=2$, $k=2$)}. The contour plots of a convolution $\Phi_\lambda(x)$ from (\ref{conv_problem})  demonstrating its multiextremal and non-smooth relief are presented in Fig.~\ref{fig1}. For estimating the Pareto set by RALR-DA, in each MCO problem 50 convolutions (\ref{conv_problem}) corresponding to 50 values $\lambda$ uniformly distributed over interval $[0,1]$ have been minimized.

@@

The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds to the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA. The averaged HV indices obtained while solving the two-dimensional bi-criteria problems are presented in \textcolor{red}{ Tab.~\ref{tab4_6} (series 1) where column $N$ is the dimension of the MCO problems, column $k$ corresponds to criteria numbers and column ``Pareto'' corresponds to the estimation obtained by the grid method.}

@@

\begin{table}
\caption{\textcolor{red}{ Results of the experiment with multidimensional MCO problems.}}
\centering
\begin{tabular}{cccccccccc} \hline
Series& $N$& $k$& Pareto & RALR-DA & OMOPSO &SMPSO & NSGAII & IBEA & SPEA2  \\ \hline
1& 2&	2 & 6.62 & 6.44 & 5.93 & 5.88 & 5.71 & 5.65 & 5.61 \\ 
2& 4&	2 & 5.04 & 4.71 & 4.45 & 4.37 & 4.39 & 4.39 & 4.53 \\ 
3& 2&	4 &  23.89 & 23.52 & 22.87 & 22.53 & 19.82 & 20.68 & 22.80 \\ \hline
\end{tabular}
\label{tab4_6}
\end{table}

The second experiment was aimed at solving bi-criteria 4-dimensional MCO problems \textcolor{red}{($N=4$, $k=2$)}. 10 test problems have been investigated and for each of them, 100 convolutions (\ref{conv_problem}) with uniformly distributed weight coefficients $\lambda$ were minimized by RALR-DA. Tab.~\ref{tab4_6} (series 2) contains the results achieved by the methods participated in the experiment.

The third collection of test problems consisted of 10 two-dimensional MCO benchmarks with 4 criteria \textcolor{red}{($N=2$, $k=4$)}. In this experiment RALR-DA in each MCO problem minimized convolutions (\ref{conv_problem}) corresponding to 100 different values of weight coefficients $\lambda$ uniformly distributed in the domain (\ref{lambda}). Like  the  previous experiments, the results are presented in Tab.~\ref{tab4_6} (series 3).

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{fig4_6.pdf} 
\caption{\textcolor{red}{ Relative backlog of the HV index ($HV_{alg} / HV_{Pareto} $). }}\label{fig4_6} 
\end{figure}

\textcolor{red}{Fig.~\ref{fig4_6} shows the ratio of the HV index of a specific algorithm to the HV index of Pareto set.}

>>>>

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
%
\title{An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria\thanks{The work was supported by the Ministry of Science and Higher Education of the Russian Federation (project no.0729-2020-0055) and by the Research and Education Mathematical Center (project no. 075-02-2020-1483/1).}}
%
\titlerunning{Algorithm for finding efficient solutions in multiobjective problems}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Konstantin Barkalov\orcidID{0000-0001-5273-2471} \and
Vladimir Grishagin\orcidID{0000-0002-2884-3670} \and
Evgeny Kozinov\orcidID{0000-0001-6776-0096}}
%
\authorrunning{K. Barkalov, V. Grishagin and E. Kozinov}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, Russia \email{\{konstantin.barkalov,evgeny.kozinov\}@itmm.unn.ru}, \email{vagris@unn.ru}}

@@

In the first experiment 100 two-dimensional bi-criteria problems have been taken ($N=2$, $k=2$). The contour plots of a convolution $\Phi_\lambda(x)$ from (\ref{conv_problem})  demonstrating its multiextremal and non-smooth relief are presented in Fig.~\ref{fig1}. For estimating the Pareto set by RALR-DA, in each MCO problem 50 convolutions (\ref{conv_problem}) corresponding to 50 values $\lambda$ uniformly distributed over interval $[0,1]$ have been minimized.

@@

The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds to the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA. The averaged HV indices obtained while solving the two-dimensional bi-criteria problems are presented in Tab.~\ref{tab4_6} (series 1) where column $N$ is the dimension of the MCO problems, column $k$ corresponds to criteria numbers and column ``Pareto'' corresponds to the estimation obtained by the grid method.

@@

\begin{table}
\caption{Results of the experiment with multidimensional MCO problems.}
\centering
\begin{tabular}{cccccccccc} \hline
Series& $N$& $k$& Pareto & RALR-DA & OMOPSO &SMPSO & NSGAII & IBEA & SPEA2  \\ \hline
1& 2&	2 & 6.62 & 6.44 & 5.93 & 5.88 & 5.71 & 5.65 & 5.61 \\ 
2& 4&	2 & 5.04 & 4.71 & 4.45 & 4.37 & 4.39 & 4.39 & 4.53 \\ 
3& 2&	4 &  23.89 & 23.52 & 22.87 & 22.53 & 19.82 & 20.68 & 22.80 \\ \hline
\end{tabular}
\label{tab4_6}
\end{table}

The second experiment was aimed at solving bi-criteria 4-dimensional MCO problems ($N=4$, $k=2$). 10 test problems have been investigated and for each of them, 100 convolutions (\ref{conv_problem}) with uniformly distributed weight coefficients $\lambda$ were minimized by RALR-DA. Tab.~\ref{tab4_6} (series 2) contains the results achieved by the methods participated in the experiment.

The third collection of test problems consisted of 10 two-dimensional MCO benchmarks with 4 criteria ($N=2$, $k=4$). In this experiment RALR-DA in each MCO problem minimized convolutions (\ref{conv_problem}) corresponding to 100 different values of weight coefficients $\lambda$ uniformly distributed in the domain (\ref{lambda}). Like  the  previous experiments, the results are presented in Tab.~\ref{tab4_6} (series 3).

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth]{fig4_6.pdf} 
\caption{Relative backlog of the HV index ($HV_{alg} / HV_{Pareto} $).}\label{fig4_6} 
\end{figure}

Fig.~\ref{fig4_6} shows the ratio of the HV index of a specific algorithm to the HV index of Pareto set.
